%\chapter{The world of Code}
\chapter{How to write code}
\label{chap:worldcode}

\begin{abstract}{Abstract}  
Programming is no longer a solitary activity, and almost all questions, problems, and error messages have been encountered and solved before. This chapter explains the most common forms of collaboration and sources of outside help, as well as giving best practices on how to write and share code yourself.
\end{abstract}

\keywords{package, library, errors, computational hygiene, notebooks}

\begin{objectives}
\item Understand the importance of re-using code when programming
\item Help first coders to avoid getting stuck
\item Explain ``computational hygiene'' and show best practices in R and Python to write and share code
\end{objectives}

In the \refchap{programmingconcepts}, you have learned how to write
your first lines of code.  You created objects of different types,
used and wrote some functions, and explored the major control structures.
You are probably eager to write your first longer piece of code and
produce some interesting data processing or analysis script. In this
chapter, we prepare you to do this with as little frustration as possible.
You will be introduced to some best practices so that you can implement
them right from the start and to some tools that make your life easier.

First, we will answer the question how you avoid reinventing
the wheel: When is it appropriate to simply use someone else's existing code, and
when do you need to write your own code from scratch? And is there a middle ground?

Second, we will discuss how to turn error messages -- which you will inevitably
see a lot -- from a frustrating annoyance into a helpful tool.

Finally, we will discuss some best practices when writing code.

\input{chapter04/reusingcode}

\input{chapter04/errors}

\input{chapter04/practices}
